---
title: "Covid91 vaccine study Final2024"
author: "Galilea Hernandez"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
male = filter(FinalData, sex =="M")
female = filter (FinalData, sex =="F")
lgbtq = filter(FinalData, LGBTQ =="gay")
druguser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

### Graphical Results


```{r}
dd2 <- male %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The graphical data shows us that male individuals that took the vaccine in contrast to the male individuals that took the placebo were less likey to get Covid19. Less than 30% of males that took the vaccine got sick from Covid19 while more than 60% of males that took the placebo did get sick from Covid19.



### Numerical

```{r}
table1 <- xtabs(~infected + treatment, data=male)
rowPerc(table1)
colPerc(table1)
```


The numerical results for male individuals shows us that the males that took the placebo were 73.80% more likely to get Covid19. The males that took the vaccine were at only 26.20% in comparison. We also see that males that did not get sick from covid whether they took the placebo or the vaccine were at 50.01% for the individuals that took the placebo and 49.99% for those that took the vaccine. We are also given ratios of the likelyhood of placebo and vaccine males in contracting the virus. For males that took the placebo vs males thatt were vaccinated it was a 3:1 ratio, so again males that took the placebo were more likely to get covid than those that took the vaccine.

From this data it is recommended that males take the Covid19 vaccine which statistically reduces the chance of getting sick.

### Inferential

```{r}
chisq.test(table1)
chisqtestGC(table1)
malefisher <- fisher.test(table1)
malefisher
```


Three tests were performed to identify if dying from covid is independent of an indivudals sex.The p-value for the first test 2.2e-16, for the second test the p-value is 0 and for the third test the p-value is also 2.2e-16. This indicates the statistical probability of dying from covid being related to the sex of the individual. The p-value of 2.2e-16 is very small indicating the probability's of males that died from the virus.


## Females


## LGBTQ



## Druggies


# Overall Results and Conclusions