---
title: "Covid91 vaccine study Final2024"
author: "Galilea Hernandez"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
male = filter(FinalData, sex =="M")
female = filter (FinalData, sex =="F")
lgbtq = filter(FinalData, LGBTQ =="gay")
druguser = filter(FinalData, DrugUser =="yes")
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males

### Graphical Results


```{r}
dd2 <- male %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The graphical data shows us that male individuals that took the vaccine in contrast to the male individuals that took the placebo were less likey to get Covid19. Less than 30% of males that took the vaccine got sick from Covid19 while more than 60% of males that took the placebo did get sick from Covid19.



### Numerical

```{r}
table1 <- xtabs(~infected + treatment, data=male)
rowPerc(table1)
colPerc(table1)
```


The numerical results for male individuals shows us that the males that took the placebo were 73.80% more likely to get Covid19. The males that took the vaccine were at only 26.20% in comparison. We also see that males that did not get sick from covid whether they took the placebo or the vaccine were at 50.01% for the individuals that took the placebo and 49.99% for those that took the vaccine. We are also given ratios of the likelyhood of placebo and vaccine males in contracting the virus. For males that took the placebo vs males thatt were vaccinated it was a 3:1 ratio, so again males that took the placebo were more likely to get covid than those that took the vaccine.

From this data it is recommended that males take the Covid19 vaccine which statistically reduces the chance of getting sick.

### Inferential

```{r}
chisq.test(table1)
chisqtestGC(table1)
malefisher <- fisher.test(table1)
malefisher
```


Three tests were performed to identify if dying from covid is independent of an individuals sex.The p-value for the first test 2.2e-16, for the second test the p-value is 0 and for the third test the p-value is also 2.2e-16. This indicates the statistical probability of dying from covid being related to the sex of the individual. The p-value of 2.2e-16 is very small indicating the probability's vaccine effectiveness in male individuals. The odds ratio represents how much more likely males that took a placebo vs the vaccine were to get the Covid19 virus.

### Confidence Interval


```{r}
effect <- malefisher$p.value/(malefisher$p.value +1)
effectivenessmale = malefisher$estimate/(malefisher$estimate +1)*100
print(effectivenessmale)
```

```{r}
low=(malefisher$conf.int[1]/(malefisher$conf.int[1] +1))*100
hi=(malefisher$conf.int[2]/(malefisher$conf.int[2] +1))*100
options(digits=2)
```



Above is the effectiveness for the covid-19 vaccine for males. If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we can conclude that the vaccine is effective for males from getting covid-19. 

The odds ratio is approximately how effective the vaccine is. The odds ratio for the guys can be seen above as 73.8. The effectiveness confidence interval for males is 68.85 to 78.32.


## Females

### Graphical Results

```{r}
dd2 <- female %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

The graphical data shows us that female individuals that took the vaccine in contrast to the female individuals that took the placebo were less likely to get Covid19. Less than 40% of females that took the vaccine got sick from Covid19 while more than 60% of females that took the placebo did get sick from Covid19. However it is important to mention that a large number of females still got sick from covid even if they took the vaccine.


### Numerical

```{r}
table1 <- xtabs(~infected + treatment, data=female)
rowPerc(table1)
colPerc(table1)
```

The numerical results for female individuals shows us that the females that took the placebo were 64.93% more likely to get Covid19. The females that took the vaccine were at 35.07% in comparison. We also see that females that did not get sick from covid whether they took the placebo or the vaccine were at 49.94% for the individuals that took the placebo and 50.06% for those that took the vaccine. We are also given ratios of the likelihood of placebo and vaccine females in contracting the virus. For females that took the placebo vs females that were vaccinated it was a 3:1 ratio, so again females that took the placebo were more likely to get covid than those that took the vaccine.

From this data we can see that while the vaccine is mostly effective, a lot of females that took the vaccine still got sick from Covid, However more half of females that took the placebo got sick, so it is still recommended that females take the vaccine.


### Inferential

```{r}
chisq.test(table1)
chisqtestGC(table1)
femalefisher <- fisher.test(table1)
femalefisher
```


Three tests were performed to identify if dying from covid is independent of an individuals sex.The p-value for the first test 1.914e-08, for the second test the p-value is 0 and for the third test the p-value is also 1.246e-08. This indicates the statistical probability of dying from covid being related to the sex of the individual. The p-values of 1.246e-08 and 1.914e-08 are very small indicating the probability's vaccine effectiveness in female individuals. The odds ratio of 1.856237 represents how much more likely females that took a placebo vs the vaccine were to get the Covid19 virus.



### Confidence Interval

```{r}
effect <- femalefisher$p.value/(femalefisher$p.value +1)
effectivenessfemale = femalefisher$estimate/(femalefisher$estimate +1)*100
print(effectivenessfemale)
```

```{r}
low=(femalefisher$conf.int[1]/(femalefisher$conf.int[1] +1))*100
hi=(femalefisher$conf.int[2]/(femalefisher$conf.int[2] +1))*100
options(digits=2)
```


Above is the effectiveness for the covid-19 vaccine for females. If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we can conclude that the vaccine is effective for females from getting covid-19. 

The odds ratio is approximately how effective the vaccine is. The odds ratio for the females can be seen above as 65. The effectiveness confidence interval for females is 59.80 to 69.92.


## LGBTQ

### Graphical Results

```{r}
dd2 <- lgbtq %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```


The graphical data shows us that lgbtq individuals that took the vaccine in contrast to the lgbtq individuals that took the placebo were less likely to get Covid19. About 20% of lgbtq individuals that took the vaccine got sick from Covid19 while more than 70% nearly 80% of lgbtq individuals that took the placebo did get sick from Covid19. The vaccine seems to be very effective for this group.


### Numerical Results

```{r}
table1 <- xtabs(~infected + treatment, data=lgbtq)
rowPerc(table1)
colPerc(table1)
```

The numerical results for lgbtq individuals shows us that the lgbtq group that took the placebo were 80% more likely to get Covid19. The lgbtq group that took the vaccine were at 20% in comparison. We also see that the lgbtq group that did not get sick from covid whether they took the placebo or the vaccine were at 48% for the individuals that took the placebo and 52% for those that took the vaccine. We are also given ratios of the likelihood of placebo and vaccine lgbtq in contracting the virus. For the lgbtq group that took the placebo vs lgbtq individuals that were vaccinated it was a 3:0 ratio, so again lgbtq's that took the placebo were more likely to get covid than those that took the vaccine.

From this data we can see that it is highly recommended that lgbtq individuals take the covid19 vaccine as statistically it is very effective.


### Inferential Results

```{r}
chisq.test(table1)
chisqtestGC(table1)
lgbtqfisher <- fisher.test(table1)
lgbtqfisher
```


Three tests were performed to identify if dying from covid is independent of an individuals sexual orientation. The p-value for the first test 2e-06, for the second test the p-value is 0 and for the third test the p-value is 7e-07. This indicates the statistical probability of dying from covid being related to the sexual orientation of the individual. The p-value of 2e-06 is very small indicating the probability's vaccine effectiveness in lgbtq individuals. The odds ratio of 4.3 represents how much more likely lgbtq's that took a placebo vs the vaccine were to get the Covid19 virus.


### Confidence Interval

```{r}
effect <- lgbtqfisher$p.value/(lgbtqfisher$p.value +1)
effectivenesslgbtq = lgbtqfisher$estimate/(lgbtqfisher$estimate +1)*100
print(effectivenesslgbtq)
```

```{r}
low=(lgbtqfisher$conf.int[1]/(lgbtqfisher$conf.int[1] +1))*100
hi=(lgbtqfisher$conf.int[2]/(lgbtqfisher$conf.int[2] +1))*100
options(digits=2)
```


Above is the effectiveness for the covid-19 vaccine for females. If the effectiveness is above 50% that means that the vaccine is effective. Based on the interval seen above, we can conclude that the vaccine is effective for lgbtq individuals from getting covid-19. 

The odds ratio is approximately how effective the vaccine is. The odds ratio for the lgbtq individuals can be seen above as 81. The effectiveness confidence interval for lgbtq's is 69 to 90.



## Druggies


# Overall Results and Conclusions